A general Nullstellensatz for generalized spaces

### 2019-03-13 paper-qcoh.tex
More work on the higher-order case

\begin{thm}\label{thm:higher-order-nullstellensatz}
Let~$\TT$ be a geometric theory. Let~$x_1\?X_1,\ldots,x_n\?X_n$ be a context over the
signature of~$\TT$. Then the canonical morphism
\[ \mathrm{Form}^\star_{\vec x}(\ul{\TT}/U_\TT)/(\dashv\vdash_{\vec x}) \longrightarrow P(X_1 \times
\cdots \times X_n) \]
sending the equivalence class of a geometric$^\star$ formula~$\varphi$ over the
signature of~$\ul{\TT}/U_\TT$ to the subset~$\{ (x_1,\ldots,x_n) \,|\, \varphi
\}$ is an isomorphism.
% XXX explain equivalence relation
\end{thm}

@@

>>>>

% XXX introduction

By \emph{geometric higher-order logic} we mean the extension of geometric
logic where we are allowed to form, in addition to the basic sorts supplied by
a given signature: finite limits of sorts; set-indexed colimits of sorts; and
powers of sorts. These derived sorts come with respective term constructors
(tuple formers, coprojections, set comprehension) and the usual rules governing
these constructors.

A \emph{geometric higher-order formula} is a formula built from equality and
relation symbols by the logical
connectives~${\top}\,{\bot}\,{\wedge}\,{\vee}\,{\exists}$
and by arbitrary set-indexed disjunctions~$\bigvee$. The existential
quantification can be over any sort, including the derived sorts. A
\emph{geometric higher-order sequent} is a sequent of the form~$(\varphi
\seq{\vec x} \psi)$ where~$\varphi$ and~$\psi$ are geometric higher-order
formulas and the sorts of the variables~$\vec x$ may be derived sorts.

The truth of geometric higher-order sequents is in general not preserved under
pullback along geometric morphisms; however for the fragment not containing
powersorts, this is true. In fact, there is a canonical way of associating to
any geometric higher-order sequent~$\sigma$ in this fragment a
family~$(\sigma_i)_i$ of ordinary geometric sequents such that~$\sigma$ is
provable in geometric higher-order logic if and only if all the
sequents~$\sigma_i$ are provable in ordinary geometric logic.

\begin{thm}\label{thm:definability}
Let~$\TT$ be a geometric theory. Let~$x_1\?X_1,\ldots,x_n\?X_n$ be a context over the
signature of~$\TT$. Then the canonical morphism
\[ \mathrm{Form}^\star_{\vec x}(\ul{\TT}/U_\TT)/(\dashv\vdash_{\vec x}) \longrightarrow P(X_1 \times
\cdots \times X_n) \]
sending the equivalence class of a geometric$^\star$ formula~$\varphi$ over the
signature of~$\ul{\TT}/U_\TT$ to the subset~$\{ (x_1,\ldots,x_n) \,|\, \varphi
\}$ is an isomorphism.
% XXX explain equivalence relation
\end{thm}

@@

\begin{proof}We argue internally to~$\Set[\TT]$.
The canonical map sends an equivalence class~$[\theta]$ to the unique map~$f :
\{(\vec x) \,|\, \varphi\} \to \{(\vec y) \,|\, \psi \}$ whose graph is given
by the set~$\{ (\vec x, \vec y) \,|\, \theta \}$.

For verifying surjectivity, let a map~$f : \{(\vec x) \,|\, \varphi\} \to
\{(\vec y) \,|\, \psi \}$ be given. Then its graph is a subset of~$\vec X
\times \vec Y$, hence by Theorem~\ref{thm:definability} given by a
geometric$^\star$ formula~$\theta$. Because~$f$ is a map, this formula is
functional; and by the Nullstellensatz, it is~$\ul{\TT}/U_\TT$-provably$^\star$ so.

For verifying injectivity, let~$\theta$ and~$\theta'$
be~$\ul{\TT}/U_\TT$-provably$^\star$ functional formulas which give rise to
identical maps. Then they also give rise to identical graphs, hence
are~$\ul{\TT}/U_\TT$-provably$^\star$ equivalent by
Theorem~\ref{thm:definability}.
\end{proof}
